---
title: "EDA cema internship task"
author: "Casey Otieno"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The dataset provided contains monthly data for children <5 years, disaggregated at a county level for the period January 2021 to June 2023.

The dataset contains the following varibales:

* Period (months from January 2021 to June 2023)
* County (the 47 counties in Kenya)
* Total number of children dewormed (Total Dewormed)
* Number of children <5 years with acute malnutrition (Acute Malnutrition)
* Number of children stunted (0-6 months, 6-23 months, 24-59 months)
* Number of children <5 years with diarrhe (Diarrhea cases)
* Number of children who are underweight (0-6 months, 6-23 months, 24-59 months)

#Install pre-requisite packages
```{r}
#install.packages("readr")
library("readr")

```

# Load and explore dataset
Load the first five rows of the dataset and summarize it.

```{r}
cema_internship_task_2023=read_csv('https://raw.githubusercontent.com/cema-uonbi/internship_task/main/data/cema_internship_task_2023.csv')
head(cema_internship_task_2023,5)
summary(cema_internship_task_2023)
dim(cema_internship_task_2023)
```

# Look for missing values and clean data
Find the total number of missing values and the number of missing values in each column if any.
```{r}
sum(is.na(cema_internship_task_2023))
colSums(is.na(cema_internship_task_2023))


#replace missing values with 0 and read the first five rows
cema_internship_task_2023[is.na(cema_internship_task_2023)]= 0
head(cema_internship_task_2023,5)
```

# Subsetting the dataset
For, example, to check the number of  children in each other variable in Kajiado county in January 23rd 2023:

```{r}
num_kaj_jan_23_2023=subset(cema_internship_task_2023, cema_internship_task_2023$county=="Kajiado County" & cema_internship_task_2023$period=='Jan-23')
num_kaj_jan_23_2023
```


## Data Visualization
# What is the total number of children dewormed in each county arranged in descending order?
```{r}
#install.packages('tidyverse')
#install.packages('dplyr')

library(tidyverse)
library(dplyr)
library(ggplot2)
library(forcats)

cema_internship_task_2023 |> 
  filter(!is.na(county)) |> 
  group_by(county) |> 
  summarize(`Total Dewormed` = sum(`Total Dewormed`, na.rm = T)) |> 
  ggplot(aes(x = fct_reorder(county, `Total Dewormed`), y = `Total Dewormed`)) +
  geom_col(width = 0.7,position =  position_dodge(0.7))+
  theme_bw()+
  labs(x="county", y='Total Dewormed',
       title = "Total number of children dewormed per county")+
    theme(text=element_text(size=7.5, face ="bold"))+
  coord_flip()
```
Nairobi county has the highest number of children dewormed as shown above.

# What is the total number of children with acute malnutrition in each county arrange in descending order?

```{r}
cema_internship_task_2023 |> 
  filter(!is.na(county)) |> 
  group_by(county) |> 
  summarize(`Acute Malnutrition` = sum(`Acute Malnutrition`, na.rm = T)) |> 
  ggplot(aes(x = fct_reorder(county, `Acute Malnutrition`), y = `Acute Malnutrition`)) +
  geom_col(width = 0.7,position =  position_dodge(0.7))+
  theme_bw()+
  labs(x="county", y='Acute Malnutrition',
       title = "Total number of children with Acute Malnutrition per county")+
    theme(text=element_text(size=7.5, face ="bold"))+
  coord_flip()
```
From above, Wajir county has the highest number of children with Acute malnutrition.

# What is the total number of children with diarrhea cases in each county arranged in descending order?

```{r}
cema_internship_task_2023 |> 
  filter(!is.na(county)) |> 
  group_by(county) |> 
  summarize(`diarrhoea cases` = sum(`diarrhoea cases`, na.rm = T)) |> 
  ggplot(aes(x = fct_reorder(county, `diarrhoea cases`), y = `diarrhoea cases`)) +
  geom_col(width = 0.7,position =  position_dodge(0.7))+
  theme_bw()+
  labs(x="county", y='diarrhoea cases',
       title = "Total number of children with diarrhoea cases per county")+
    theme(text=element_text(size=7.5, face ="bold"))+
  coord_flip()
```
From above, Nairobi is leading with the highest number of children with diarrhoea cases.

# Number of children <5 years who are underweight by age group.
```{r}
underweight_over_time <- cema_internship_task_2023 %>%
 group_by(county) %>%
 summarize(
 Underweight_0_6_months=sum(`Underweight 0-<6 months`),
 Underweight_6_23_months=sum(`Underweight 6-23 months`),
 Underweight_24_59_months=sum(`Underweight 24-59 Months`)
)
 ggplot(data= underweight_over_time ,aes(x = fct_reorder(county, `Underweight_0_6_months`), y = `Underweight_0_6_months`)) +
  geom_col(width = 0.7,position =  position_dodge(0.7))+
  theme_bw()+
  labs(x="county", y='Underweight_0_6_months',
       title = "Total number of children Underweight_0_6_months per county")+
    theme(text=element_text(size=7.5, face ="bold"))+
  coord_flip() 
 
  ggplot(data= underweight_over_time ,aes(x = fct_reorder(county, `Underweight_6_23_months`), y = `Underweight_6_23_months`)) +
  geom_col(width = 0.7,position =  position_dodge(0.7))+
  theme_bw()+
  labs(x="county", y='Underweight_6_23_months',
       title = "Total number of children Underweight_6_23_months per county")+
    theme(text=element_text(size=7.5, face ="bold"))+
  coord_flip()
  
 ggplot(data= underweight_over_time ,aes(x = fct_reorder(county, `Underweight_24_59_months`), y = `Underweight_24_59_months`)) +
  geom_col(width = 0.7,position =  position_dodge(0.7))+
  theme_bw()+
  labs(x="county", y='Underweight_24_59_months',
       title = "Total number of children Underweight_24_59_months per county")+
    theme(text=element_text(size=7.5, face ="bold"))+
  coord_flip()
 
```
the graph shows that Nairobi and Turkana county have the highest number of children underweight between 0-6, 6-23, 24-59 months while Nyamira County has the least.

 Based on the exploratory data analysis above, a research question could be: Is there a relationship between the number of children dewormed and acute malnutrition  for each month. As below, the correlation coefficient is less than 0 indicating a weak linear relationship between the total number of children dewormed and prevalance of acute malnutrition.
```{r}

# Correlation analysis - Correlation between Total Dewormed and Acute Malnutrition
correlation <- cor(cema_internship_task_2023$`Total Dewormed`, cema_internship_task_2023$`Acute Malnutrition`, use = "pairwise.complete.obs")
print(paste("Correlation between Total Dewormed and Acute Malnutrition:", correlation))

```


# Another research question: 
How have the rates of diarrhea cases and underweight in children under 5 years evolved over the period from January 2021 to June 2023 across different counties in Kenya?

First convert period column to date format
```{r}

library(lubridate)
char_dates=cema_internship_task_2023$period
cema_internship_task_2023$period <- as.Date(paste0("01-", char_dates), format= "%d-%b-%y")

class(cema_internship_task_2023$period)
head(cema_internship_task_2023,3)
```


```{r}
# Group the data by Period and calculate the sum of Diarrhea cases for each month
diarrhea_over_time <- cema_internship_task_2023 %>%
  group_by(period) %>%
  summarize(Diarrhea_cases = sum(`diarrhoea cases`))

# Line plot of Diarrhea cases over time
ggplot(diarrhea_over_time, aes(x = period, y = Diarrhea_cases)) +
  geom_line() +
  theme_bw()+
  labs(x = "Period", y = "Diarrhea Cases", title = "Number of Children <5 Years with Diarrhea Over Time")
```
The graph above shows an uptrend, the number of children with diarrhoea cases has increased overtime.
```{r}

# Group the data by Period and calculate the sum of Underweight cases for each age group and month
underweight_over_time <- cema_internship_task_2023 %>%
  group_by(period) %>%
  summarize(
    Underweight_0_6_months = sum(`Underweight 0-<6 months`),
    Underweight_6_23_months = sum(`Underweight 6-23 months`),
    Underweight_24_59_months = sum(`Underweight 24-59 Months`)
  )

# Line plot of Underweight cases by age group over time
ggplot(underweight_over_time, aes(x = period)) +
  geom_line(aes(y = Underweight_0_6_months, color = "0-6 months")) +
  geom_line(aes(y = Underweight_6_23_months, color = "6-23 months")) +
  geom_line(aes(y = Underweight_24_59_months, color = "24-59 months")) +
  theme_bw()+
  labs(x = "Period", y = "Underweight Cases", title = "Number of Children Who Are Underweight by Age Group Over Time") +
  scale_color_manual(values = c("red", "blue", "green"), name = "Age Group")
```
From the graph above, children between the age of 6-23 months have recorded the highest number of underweight children among the other age groups, and has gradually increased over time.Children between 0-6 have less underweight cases.




